Carryover effects in free recall reveal how past experiences influence memories of future experiences

### 2023-01-16 paper/main.tex
backing up progress on discussion section   - added a more detailed summary and overview of the key results

We sought to manipulate two aspects of how participants memorized sequences of
word lists. First, we added two additional sources of visual variation to the
individual word presentations: font color and onscreen location. Importantly,
these visual features were independent of the meaning or semantic content of
the words (e.g., word category, size of the refferent) and of the lexicographic
properties of the word (e.g., word length, first letter). We wondered whether
this additional word-independent information might facilitate recall (e.g., by
providing new potential ways of organizing or retrieving memories of the
studied words) or impair recall (e.g., by distracting participants). Second,
our primary experimental manipulations entailed manipulating the orders in
which words were studied (and how those orderings changed over time). We
wondered whether presenting the same list of words in different orders (e.g.,
sorted along one feature dimension versus another) might serve to influence how
participants organized their memories of the words. We also wondered whether
some order manipulations might be temporally ``sticky'' by influencing how
\textit{future} lists were remembered.

To obtain a clean preliminary estimate of the consequences on memory of
randomly varying the font colors and locations of presented words (versus
holding the font color fixed at black, and holding the display locations fixed
at the center of the display) we compared participants' performance on the
\textit{feature rich} and \textit{reduced} experimental conditions (see
\textit{Random conditions}, Fig.~\dynamicsRandom). In the feature rich
condition the words' colors and locations varied randomly across words, and in
the reduced condition words were always presented in black, at the center of
the display. Aggregating across all lists for each participant, we found no
difference in recall accuracy for feature rich versus reduced lists ($t(126) =
-0.290, p = 0.772$). However, participants in the feature rich condition
clustered their recalls substantially more along every dimension we examined
(temporal clustering: $t(126) = 10.624, p < 0.001$; category clustering:
$t(126) = 10.077, p < 0.001$; size clustering: $t(126) = 11.829, p < 0.001$;
word length clustering: $t(126) = 10.639, p < 0.001$; first letter clustering:
$t(126) = 7.775, p = 0.000$; see \textit{Permutation-corrected feature
clustering scores} for more information about how we quantified each
participant's clustering tendencies.) Taken together, these comparisons suggest
that adding new features changes how participants organize their memories of
studied words, even when those new features are independent of the words
themselves and even when the new features vary randomly across words. We found
no evidence that those additional uninformative features were distracting (in
terms of their impact on memory performance), but they did affect participants'
recall dynamics (measured via their clustering scores).

@@

We asked participants to study and freely recall word lists. The words on each
list (and the set of lists) were held constant across participants. For each
word, we considered (and manipulated) two semantic features (category and size)
that reflected aspects of the \textit{meanings} of the words, along with two
lexicographic features (word length and first letter), which reflected aspects
of the words' \textit{letters}. These semantic and lexicographic features are
intrinsic to each word. We also considered and manipulated two additional
visual features (color and location) that affected the \textit{appearance} of
each studied item, but could be varied independently of the words' identities.
Across different experimental conditions, we manipulated how the visual
features varied across words (within each list), along with the orders of each
list's words. Although participants' task (verbally recalling as many words as
possible, in any order, within one minute) remained constant across all of
these conditions, and although the set of words they studied on each list
remained constant, our manipulations substantially affected participants'
memories. The impact of some of the manipulations also affected how
participants remembered \textit{future} lists that were sorted randomly.

@@

>>>>

While holding the set of words (and the assignments of words to lists)
constant, we manipulated two aspects of participants' experiences of studying
each list. We sought to understand the effects of these manipulations on
participants' memories for the studied words. First, we added two additional
sources of visual variation to the individual word presentations: font color
and onscreen location. Importantly, these visual features were independent of
the meaning or semantic content of the words (e.g., word category, size of the
refferent, etc.) and of the lexicographic properties of the words (e.g., word length,
first letter, etc.). We wondered whether this additional word-independent information
might facilitate recall (e.g., by providing new potential ways of organizing or
retrieving memories of the studied words) or impair recall (e.g., by
distracting participants with irrelevant information). Second, we manipulated
the orders in which words were studied (and how those orderings changed over
time). We wondered whether presenting the same list of words with different
appearances (e.g., by manipulating font size and onscreen location) or in
different orders (e.g., sorted along one feature dimension versus another)
might serve to influence how participants organized their memories of the
words. We also wondered whether some order manipulations might be temporally
``sticky'' by influencing how \textit{future} lists were remembered.

To obtain a clean preliminary estimate of the consequences on memory of
randomly varying the font colors and locations of presented words (versus
holding the font color fixed at black, and holding the display locations fixed
at the center of the display) we compared participants' performance on the
\textit{feature rich} and \textit{reduced} experimental conditions (see
\textit{Random conditions}, Fig.~\dynamicsRandom). In the feature rich
condition the words' colors and locations varied randomly across words, and in
the reduced condition words were always presented in black, at the center of
the display. Aggregating across all lists for each participant, we found no
difference in recall accuracy for feature rich versus reduced lists ($t(126) =
-0.290, p = 0.772$). However, participants in the feature rich condition
clustered their recalls substantially more along every dimension we examined
(temporal clustering: $t(126) = 10.624, p < 0.001$; category clustering:
$t(126) = 10.077, p < 0.001$; size clustering: $t(126) = 11.829, p < 0.001$;
word length clustering: $t(126) = 10.639, p < 0.001$; first letter clustering:
$t(126) = 7.775, p < 0.001$; see \textit{Permutation-corrected feature
clustering scores} for more information about how we quantified each
participant's clustering tendencies.) Taken together, these comparisons suggest
that adding new features changes how participants organize their memories of
studied words, even when those new features are independent of the words
themselves and even when the new features vary randomly across words. We found
no evidence that those additional uninformative features were distracting (in
terms of their impact on memory performance), but they did affect participants'
recall dynamics (measured via their clustering scores).

@@

We asked participants to study and freely recall word lists. The words on each
list (and the total set of lists) were held constant across participants. For
each word, we considered (and manipulated) two semantic features (category and
size) that reflected aspects of the \textit{meanings} of the words, along with
two lexicographic features (word length and first letter), which reflected
aspects of the words' \textit{letters}. These semantic and lexicographic
features are intrinsic to each word. We also considered and manipulated two
additional visual features (color and location) that affected the
\textit{appearance} of each studied item, but could be varied independently of
the words' identities. Across different experimental conditions, we manipulated
how the visual features varied across words (within each list), along with the
orders of each list's words. Although participants' task (verbally recalling as
many words as possible, in any order, within one minute) remained constant
across all of these conditions, and although the set of words they studied on
each list remained constant, our manipulations substantially affected
participants' memories. The impact of some of the manipulations also affected
how participants remembered \textit{future} lists that were sorted randomly.

@@

We found that participants in our feature rich condition (where we varied
words' appearances) recalled similar proportions of words to participants in a
reduced condition (where appearance was held constant across words). However,
varying the words' appearances led participants to exhibit much more temporal
and feature-based clustering. This suggests that even seemingly irrelevant
elements of our experiences can affect how we remember them.

When we held the within-list variability in participants' visual experiences
fixed across lists (in the feature rich and reduced conditions), they
remembered more words on early versus late lists. On feature rich lists, they
also showed stronger clustering on early versus late lists. However, when we
\textit{varied} participants' visual experiences across lists (in the ``reduced
(early)'' and ``reduced (late)'' conditions), these early versus late accuracy
and clustering differences disappeared. Abruptly changing how irrelevant visual
features change across words seems to act as a sort of ``event boundary'' that
partially resets how participants process and remember post-boundary lists.
Within-list clustering also increases in these manipulations, suggesting that
the "within-event" words are being more tightly associated with each other.

When we held the visual features constant on early lists, but then varied
words' appearances on later lists (i.e., the reduced (early) condition), this improved
participants' overall memory performance. However, this impact was directional:
when we \textit{removed} visual features on late lists that had been present on
early lists (i.e., the reduced (late) condition), we saw no memory improvement.

When we (stochastically) sorted early lists along different feature dimensions,
we found several impacts on participants' memories. Sorting early lists
semantically (by word category) enhanced participants' memories for those
lists, but the effects on performance of sorting along other feature dimensions
were inconclusive. However, each order manipulation substantially affected how
participants \textit{organized} their memories of words from the ordered lists.
When we sorted lists semantically participants displayed stronger semantic
clustering; when we sorted lists lexicographically they displayed stronger
lexicographic clustering; and when we sorted lists visually they displayed
stronger visual clustering. Clustering along the unmanipulated feature
dimensions in each of these cases was unchanged.

The order manipulations we examined also appeared to induce, in some cases, a
tendancy to ``clump'' similar words within a list. This was most apparent on
semantically ordered lists, where the probability of initiating recall with a
given word seemed to follow groupings defined by feature changepoints.

We also examined the impact of early list order manipulations on memory for
late lists. At the group level, we found little evidence for lingering
``carryover'' effects of these manipulations; participants in the order
manipulation conditions showed similar memory performance and clustering on
late lists to participants in the corresponding control (feature rich)
condition. At the level of individual participants, however, we found several
meaningful patterns.

Participants who showed stronger feature clustering on early (order
manipulated) lists tended to better remember late (randomly ordered) lists.
Participants who remembered early lists better also tended to show stronger
feature clustering (along their condition's feature dimension) on late lists
(even though the words on those late lists were presented in a random order).
We also observed some (weaker) carryover effects of temporal clustering.
Participants who showed stronger feature clustering (along their condition's
feature dimension) on early lists tended to show stronger temporal clustering
on late lists. And participants who showed stronger temporal clustering on
early lists also tended to show stronger feature clustering on late lists.
Essentially, these order manipulations appeared to affect each participant
differently. Some participants were sensitive to our manipulations, and those
particpants showed stronger impacts on their memory performance for the ordered
lists as well as future (random) lists. Other participants appeared relatively
insensitive to our manipulations, and those participants showed little
carryover effects on late lists.

These results at the individual participant level suggested to us that either
(a) some participants were more sensitive to \textit{any} order manipulation,
or (b) some participants might be more (or less) sensitive to manipulations
along \textit{particular} (e.g., preferred) feature dimensions. To help
distinguish between these possibilities, we designed an adaptive manipulation
whereby we attempted to manipulate whether participants studied words in an
order that matched (or mismatched) our estimate of how they would cluster or
organize the studied words in memory (i.e., their idiosyncratic memory
fingerprint). We found that when we presented words in orders that were
consistent with participants' memory fingerprints, they remembered more words
overall and showed stronger temporal clustering. This comports well with the
second possibility described above. Specifically, each participant seems to
bring into the experiment their own idiosynchratic preferences and strategies
for organizing the words in their memories. When we presented the words in an
order consistent with each participant's idiosyncratic strategies, their memory
performance improved. This might indicate that the participants were spending
less cognitive effort ``reorganizing'' the incoming words on those lists, which
freed up resources to devote to encoding processes instead.
